DynaProg for Scala

### 2013-01-21 docs/report/benchmarks.tex

% ----------------------------------------------
% MatrixMult-512, Mac+JDK7\ul
% Original: 24.937 sec / 24.658 sec
% Optimized concatenation: 19.329 sec / 18.924 sec
% Tabulation as arrays+inline: 14.976 sec / 14.849 sec
\newpage
% ------------------------------------------------------------------------------------------------
{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
% ------------------------------------------------------------------------------------------------
\subsection{Matrix chain multiplication}
Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.  Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is because 64-bit mode is enabled. Because external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby reducing the performance of CUDA operations.
% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
% GPU: GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?

\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
& \hdr{DynaProg}{Scala version}
	& 0.247		& 2.048		& 16.657		& 134.207	& 1101.134	&  			& 		 \\
& \hdr{Optimized}{C, single thread}
	& 0.001		& 0.008		& 0.078		& 1.179		& 19.811		& 206.556	& 2010.485 \\
& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{ADP Fusion}{\cite{adp_fusion}}
	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad\quad$}}
& \hdr{DynaProg}{CUDA version}
	& 0.028		& 0.046		& 0.129		& 0.348		& 1.685		& 10.308		& 71.215 \\
& \hdr{Optimized}{CUDA, 64-bit}
	& 0.011		& 0.020		& 0.074		& 0.315		& 1.651		& 10.354		& 72.376 \\
& \hdr{ATLP}{\cite{gpu_atlp}, rescaled results}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{GAPC}{\cite{gapc_thesis}, OpenCL}
	& 			& 			& 			& 			& 			& 			& 		 \\
\\[-10pt] \bottomrule\end{tabular}}\end{center}
\caption{Running time of matrix chain multiplication (in seconds)}\end{table}

@@

For DynaProg/Scala we use a variant of the problem description: the original version only stores the matrix multiplication score whereas the modified version also stores the matrix dimension. This allows a speedup of $2.9\times$ probably due to additional function calls to retrieve the values. Also with the default JVM parameters, the program cannot address sequences longer than $\sim 420$ elements due to a stack overflow, for these benchmarks, we increased this limit.

% 1024=0.316, 2048=1.427, 4096=8.841, 8192=62.064

@@

\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
& \hdr{DynaProg}{Scala version}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{Optimized}{C, single thread}
	& <0.001		& <0.001		& 0.002		& 0.009		& 0.042		& 0.171		& 0.706 \\
& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{ADP Fusion}{\cite{adp_fusion}}
	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad$}}
& \hdr{DynaProg}{CUDA version}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{Optimized}{CUDA, 64-bit}
	& 0.002		& 0.004		& 0.004		& 0.007		& 0.024		& 0.070		& 0.270 \\
& \hdr{CUDAlign}{\cite{swat_linear}, version 2.0}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{GAPC}{\cite{gapc_thesis}, OpenCL}
	& 			& 			& 			& 			& 			& 			& 		 \\
\\[-10pt] \bottomrule\end{tabular}}\end{center}
\caption{Running time of Smith-Waterman (in seconds)}\end{table}

>>>>

% ----------------------------------------------
% MatrixMult-512, Mac+JDK7\ul
% Original: 24.937 sec / 24.658 sec
% Optimized concatenation: 19.329 sec / 18.924 sec
% Tabulation as arrays+inline: 14.976 sec / 14.849 sec
\newpage
\subsection{Matrix chain multiplication}
Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.  Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is because 64-bit mode is enabled. Because external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby reducing the performance of CUDA operations.
% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
% GPU: GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?

\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
& \hdr{DynaProg}{Scala version}
	& 0.247		& 2.048		& 16.657		& 134.207	& 1101.134	&  			& 		 \\
& \hdr{Optimized}{C, single thread}
	& 0.001		& 0.008		& 0.078		& 1.179		& 19.811		& 206.556	& 2010.485 \\
& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{ADP Fusion}{\cite{adp_fusion}}
	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad\quad$}}
& \hdr{DynaProg}{CUDA version}
	& 0.028		& 0.046		& 0.129		& 0.348		& 1.685		& 10.308		& 71.215 \\
& \hdr{Optimized}{CUDA, 64-bit}
	& 0.011		& 0.020		& 0.074		& 0.315		& 1.651		& 10.354		& 72.376 \\
% 32-bit &&		&			&			& 0.316		& 1.427		& 8.841		& 62.064
& \hdr{ATLP}{\cite{gpu_atlp}, rescaled results$^{(1)}$}
	& 0.170		& 0.199		& 0.227		& 0.397		& 1.335		& 7.299 		& - \\
\\[-10pt] \bottomrule\end{tabular}}\end{center}
\caption{Running time of matrix chain multiplication (in seconds)}\end{table}

$^{(1)}$ Assuming that 72\% of the running time is due to memory accesses, and considering a $3.55\times$ memory throughput slowdown (see \S\ref{results_discussion}). % 0.06 & 0.07 & 0.08 & 0.14 & 0.47 & 2.57 => f[x_] := (x*(1 - .72)) + (x*0.72)*(102.4/28.8)
% 0.1704 & 0.1988 & 0.2272 & .3976 & 1.3348 & 7.2988

@@

For DynaProg/Scala we use a variant of the problem description: the original version only stores the matrix multiplication score whereas the modified version also stores the matrix dimension. This allows a speedup of $2.9\times$ probably due to additional function calls to retrieve the values. Also with the default JVM parameters, the program cannot address sequences longer than $\sim 420$ elements due to a stack overflow, for these benchmarks, we increased this limit	.

% ------------------------------------------------------------------------------------------------
{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
% ------------------------------------------------------------------------------------------------

@@

\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
& \hdr{DynaProg}{Scala version}
	& 0.115		& 0.438		& 1.743		& 7.864		& 32.788		& 139.309	& $^{(1)}\infty$ \\
& \hdr{Optimized}{C, single thread}
	& <0.001		& <0.001		& 0.002		& 0.009		& 0.042		& 0.171		& 0.706 \\
& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
	& 			& 			& 			& 			& 			& 			& 		 \\
& \hdr{ADP Fusion}{\cite{adp_fusion}}
	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad$}}
& \hdr{DynaProg}{CUDA version}
	& 0.031		& 0.037		& 0.047		& 0.068		& 0.116		& 0.202		& $^{(2)}$3.226 \\
& \hdr{Optimized}{CUDA, 64-bit}
	& 0.002		& 0.004		& 0.004		& 0.007		& 0.024		& 0.070		& 0.270 \\
& \hdr{CUDAlign}{\cite{swat_linear}, version 2.0}
	& 			& 			& 			& 			& 			& 			& 		 \\
\\[-10pt] \bottomrule\end{tabular}}\end{center}
\caption{Running time of Smith-Waterman (in seconds)}\end{table}

$^{(1)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx8G -Xms8G -XX:MaxPermSize=8G}) this emphasizes that the top-down parsing approach might not be the best for such simple problems.

$^{(2)}$ Since the memory requirements are larger than the device capacity, the backtrack matrix overflows in the main memory, thereby significantly degrading the performance ($4\times$ slowdown). This extra memory requirement is due to the use of 3 matrices to avoid non-serial dependencies (hence requiring at least $3 \cdot 2$ bytes of memory per matrix element for backtrack).

X: java.lang.OutOfMemoryError: GC overhead limit exceeded

### 2013-01-21 docs/report/implementation.tex

% ----------------------------------------------
\subsubsection{Results discussion}\ul
\item \textbf{User interface:} It has been put in evidence in \cite{perfeval_gpu} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M). With the newer architecture, this difference has been reduced to less than 3.5\%, decoupled UI and CUDA   performing best. So we can safely ignore this issue.
\item \textbf{Blocks synchronization:}\ul
	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but results still remains valid, this confirms the observation made by \cite{gpu_barrier}. Speedup for matrix size of 1024 are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
	\item In the parallelogram case, using all threads to monitor other blocks status instead of the first one only results in a 6.4x speedup (22.72$\to$3.52ms) for the parallelogram.
	\ule
\item \textbf{Multiple threads per matrix cell:} in the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increases (there is one more dependency). According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity. However, in our setup (memory layout+algorithm+hardware), we did not find any improvement by doing so. We want to explore the reason for that: we pose as hypothesis that the bandwidth is the bottleneck of our setup and test it.\ul
\item

First we need to prove that we use almost all the available memory bandwidth: for matrix multiplication, in a triangular matrix, we have
\[\text{Total transfer}=\frac{n(n+1)}{2} \text{ writes} + \sum_{i=0}^{n-1} 2 i \cdot (n-i) \text{ reads}\]
where each write is 10 bytes (long+short), and each read is 8 bytes (long). For $n=4096$ we transfer
% n*(n + 1)/2*12 + Sum[2*i*(n - i), {i, 0, n - 1}]*8
183'352'614'912 bytes which corresponds to 183.35GB. In 8.841 seconds, we can transfer theoretically at most $8.841\cdot 28.8 = 254 \rm GB$. Hence  72\% of the algorithm running time is spent into memory accesses.

\item On a 4096 matrix, if we assume that the \cite{gpu_atlp} card would have the same bandwidth as our card, their running time would be
\[2.57 \cdot (1-.72) + 2.57 \cdot 0.72 \cdot \tfrac{102.4_{GB/s}}{28.8_{GB/s}} = 9.43\rm s_{\text{ ATLP}} > 8.84\rm s_{\text{ our}}\]
This shows that our algorithm is comparable to theirs. However, we must avoid a close comparison because the fundamental hardware differences would make a tight computation almost intractable (additionally, we do not have \cite{gpu_atlp} source code).
\ule
As a conclusion, (1) it seems that the technique used in \cite{gpu_atlp} only applies to legacy hardware, however this remains a supposition (as we can not compare) and (2) we are on par, if not better with one of the best current implementations.

>>>>

% ----------------------------------------------
\subsubsection{Results discussion} \label{results_discussion} \ul
\item \textbf{User interface:} It has been put in evidence in \cite{perfeval_gpu} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M). With the newer architecture, this difference has been reduced to less than 3.5\%, decoupled UI and CUDA   performing best. So we can safely ignore this issue.
\item \textbf{Blocks synchronization:}\ul
	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but results still remains valid, this confirms the observation made by \cite{gpu_barrier}. Speedup for matrix size of 1024 are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
	\item In the parallelogram case, using all threads to monitor other blocks status instead of the first one only results in a 6.4x speedup (22.72$\to$3.52ms) for the parallelogram.
	\ule
\item \textbf{Multiple threads per matrix cell:} in the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increases (there is one more dependency). According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity. However, in our setup (memory layout+algorithm+hardware), we did not find any improvement by doing so. We want to explore the reason for that: we pose as hypothesis that the bandwidth is the bottleneck of our setup and test it.\ul
\item

First we need to prove that we use almost all the available memory bandwidth: for matrix multiplication, in a triangular matrix, we have
\[\text{Total transfer}=\frac{n(n+1)}{2} \text{ writes} + \sum_{i=0}^{n-1} 2 i \cdot (n-i) \text{ reads}\]
where each write is 10 bytes (long+short), and each read is 8 bytes (long). For $n=4096$ we transfer
% n*(n + 1)/2*12 + Sum[2*i*(n - i), {i, 0, n - 1}]*8
183'352'614'912 bytes which corresponds to 183.35GB. In 8.841 seconds, we can transfer theoretically at most $8.841\cdot 28.8 = 254.6 \rm GB$. Hence  72\% of the algorithm running time is spent into memory accesses.

\item On a 4096 matrix, if we assume that the \cite{gpu_atlp} card would have the same bandwidth as our card, their running time would be
\[2.57 \cdot (1-.72) + 2.57 \cdot 0.72 \cdot \tfrac{102.4_{GB/s}}{28.8_{GB/s}} = 7.30\rm s_{\text{ ATLP}} < 8.84\rm s_{\text{ our}}\]
This shows that our algorithm is comparable to theirs. However, we must avoid a close comparison because the fundamental hardware differences would make a tight computation almost intractable (additionally, we do not have \cite{gpu_atlp} source code).
\ule
As a conclusion, (1) it seems that the technique used in \cite{gpu_atlp} brings more performance improvement with legacy hardware, however this remains a supposition (as we can not compare) and (2) we are slightly worse than one of the best current implementations.
